ThermoMaze: A behavioral paradigm for readout of immobility-related brain events

Version 1

Abstract

Abstract

Brain states fluctuate between exploratory and consummatory phases of behavior. These state changes affect both internal computation and the organism’s responses to sensory inputs. Understanding neuronal mechanisms supporting exploratory and consummatory states and their switching requires experimental control of behavioral shifts and collecting sufficient amounts of brain data. To achieve this goal, we developed the ThermoMaze, which exploits the animal’s natural warmth-seeking homeostatic behavior. By decreasing the floor temperature and selectively heating unmarked areas, mice avoid the aversive state by exploring the maze and finding the warm spot. In its design, the ThermoMaze is analogous to the widely used water maze but without the inconvenience of a wet environment and, therefore, allows the collection of physiological data in many trials. We combined the ThermoMaze with electrophysiology recording, and report that spiking activity of hippocampal CA1 neurons during sharp-wave ripple events encode the position of the animal. Thus, place-specific firing is not confined to locomotion and associated theta oscillations but persist during waking immobility and sleep at the same location. The ThermoMaze will allow for detailed studies of brain correlates of immobility, preparatory-consummatory transitions and open new options for studying behavior-mediated temperature homeostasis.

Introduction

All behaviors can be considered as parts of a sequence of action-rest transition1. Brain states in vertebrates fall into dichotomous categories, and correspond roughly to what early behavioral research referred to as “preparative” (or “exploratory”) and “consummatory” (or “terminal”) classes2. In mammals, these two fundamental brain states can be readily identified by basic electrophysiological monitoring of various brain structures3. They are also referred to as voluntary and non-voluntary or conscious and non-conscious brain states3. Switching between these states is correlated with high and low release of subcortical neuromodulators4–9. Consummatory behaviors include feeding and drinking, resting and its extreme form, non-rapid eye movement (NREM) sleep. Preparatory and consummatory behaviors in the hippocampus are associated with theta oscillations and sharp wave ripples (SPW-Rs), respectively10.

Deciphering the physiological underpinnings of these categories and revealing the significance of brain state transitions for cognition requires sufficient sampling of the relevant brain states. This is usually achieved by extended repeated recordings or, when possible, recording large numbers of neurons simultaneously. Prolongation of explorative behavior can be readily achieved by placing the animal in novel environments, by food or water deprivation or introducing delays in choice behavior tasks11,12. Recently, the honeycomb maze paradigm was introduced to extend the observation periods of explorative deliberation13.

In contrast, the experimental control of consummatory classes of behavior is more difficult. Sleep provides an opportunity for long recordings. Comparison of sleep before and after learning is a standard paradigm to examine experience-induced brain plasticity14,15. Consummatory brain states associated with eating, drinking and sex change rapidly with satiety and requires prolonged periods of deprivation16–19. Controlling periods of awake immobility is most difficult20–22, mainly because forced immobilization of the animal is stressful23 and is accompanied by altered physiological states24.

Here we introduce the ThermoMaze, a behavioral paradigm that allows for the collection of large amounts of physiological data while the animal rests at distinct experimenter-controlled locations. In standard laboratory environments (20-24 °C)25, both housing and data collection take place below the thermoneutral zone of mice (26-34 °C)26–28. The ThermoMaze exploits the animal’s behavioral thermoregulation mechanisms29,30 and promotes thermotaxis (i.e., movement in response to environmental temperature)31. Searching for a warmer environment, social crowding and nest building are natural behavioral components of heat homeostasis31–33. The ThermoMaze allows the experimenter to guide small rodents to multiple positions in a two-dimensional environment. Decreasing the maze floor temperature induces heat seeking behavior and after finding a warm spot, the animal stays immobile at that spot for extended periods of time, allowing for recording large amounts of neurophysiological data in immobility-related brain states. We report on both behavioral control and hippocampal electrophysiological correlates of heat seeking activity to illustrate the versatile utility of the ThermoMaze.

Results

Design and Construction of the ThermoMaze

The ThermoMaze is designed to guide small rodents to warm spatial locations in a two-dimensional cold environment, consisting of a box (width, length, height: 20, 20, 40 cm, respectively) made from an acrylic plexiglass sheet (Fig. 1A, top). The floor of the maze is constructed from 25 Peltier elements (40 x 40 x 3.6 mm) that are attached to aluminum water cooling block heatsinks (40 x 40 x 12 mm, n = 25) with heat-conductive epoxy and are insulated from each other by wood epoxy (Fig. 1A, dashed inset). Each Peltier element is controlled by an electrically operated switch (relay) that opens and closes high-current circuits by receiving transistor-transistor-logic (TTL) signals from outside sources (Fig. 1B). Peltier elements can be heated individually up to 30 °C to provide a warm spot for the animal when other regions of the floor are under cooling (Fig. 1B, active heating of one Peltier element is shown). The ambient temperature of the maze is controlled by water circulated from the water tank through the water-cooling blocks. We set the floor temperature to either ∼25°C (room temperature) or to ∼10 °C (cooling, Fig. 1C and Suppl. Fig. 1), but a range of ambient temperatures (5-30 °C) could be employed. The water temperature is monitored by a K-type thermocouple placed inside the water tank (Fig. 1A bottom). The floor temperature of the ThermoMaze is monitored using a thermal camera (FLIR C5) providing continuous registration of real-time temperature changes (Fig. 1A).

Construction and temperature control of the ThermoMaze.

A) Schematic of the ThermoMaze. The floor was built using 25 Peltier elements attached to water cooling block heatsinks (building block). The position of the animal and the temperature of the ThermoMaze can be recorded using a video camera and an infrared camera positioned above the box, respectively. An ‘X’ was taped inside the maze as an external cue below the camera synchronizing LED. Water circulates through the water cooling heatsinks using a water pump submerged in a water tank (one row of heatsinks is attached to one pump). The temperature of the water tank is monitored and recorded using a thermocouple (white symbol inside water tank, DAQ – analog input of the data acquisition system). Peltier elements are connected to a power supply (red and blue dots represent the anode and cathode connection). B) Circuit diagram and schematic of Peltier elements (n = 25), viewed from the top. TTL pulses generated by an AVR-based microcontroller board (Arduino Mega 2560) close a relay switch connected to a variable voltage power source. Each Peltier element can be independently heated (surface temperature depends on applied voltage and temperature difference between hot and cold plate of Peltier element). C) Schematic of the water circulation cooling system, viewed from the bottom of the floor (each Peltier element has its own water-cooling aluminum heatsink, shown in silver, n = 25). Five submerging DC pumps are used to circulate water across 25 heatsinks (dashed lines show the Peltier elements connected to one pump). The temperature of the heatsink is transferred to the Peltier element passively through the silver epoxy resulting in passive cooling of the floor of the ThermoMaze.

Prior to the experiments, the thermal camera, which continuously measures the surface temperature of the floor of the ThermoMaze is calibrated by thermocouples placed directly on Peltier elements (Fig. 2). The accuracy of the FLIR 5C infrared camera is ±3 °C. With proper calibration and attention to emissivity (an object’s ability to emit rather than reflect infrared energy) the margin of error can be less than 1 °C34.

Calibration of the ThermoMaze temperature regulation.

A) Side view of the ThermoMaze. Prior to animal experiments, we calibrated the heating and cooling performance of the Peltier elements and temperature measurement. We attached thermocouples (white symbols) to the surface of the Peltier elements serving as the ground-truth for calibrating the infrared camera placed above the ThermoMaze. Different voltage levels were used for the calibration (2.2, 2.4, 2.6, 2.8 and 3V) while the water tank temperature was kept constant. B) Top: four Peltier elements used in later experiments are chosen for calibration (four corners). Bottom: one minute heating was repeated four times at each voltage level. C) Simultaneously recorded temperature by thermocouples (left) and infrared camera (right). Increasing voltages induced increased heating (n = 4 trials per intensity, mean ± SD are shown). While the temporal dynamics yielded similar results between the two systems, we found ∼4 °C offset between infrared and thermocouple-measured signals. D) Temperature changes of four Peltier elements used during an emulated behavioral session (without any animal subject) tracked by thermocouples. E) Temporal dynamics of temperature changes at the four Peltier elements during active heating and following passive cooling. The temperature reaches steady state within 31 ± 10.3 seconds (mean ± SD, n = 4 trials across 4 Peltier elements).

Mice seek out hidden warm spots in the ThermoMaze

To illustrate the novel advantages of the ThermoMaze on behavior and brain activity, we tested 11 mice (n = 3 male and 8 female mice) with silicon probe recordings from the hippocampus (Suppl. Table 1). One wall was marked by a prominent visual cue (black tape and blinking light-emitting diode; LED) to provide a distinct spatial cue in the box (Fig. 1A)35. On each experimental day, the mouse was placed in the ThermoMaze and allowed to explore it for 10 min at room temperature (“Pre-cooling” sub-session; Figure 3A). Next, the ThermoMaze temperature was decreased to around 14 °C for 80 min and four Peltier elements (“warm spots”; typically, in the corners) were sequentially and repeatedly turned on and heated up to 30 °C. One Peltier element was turned on for 5 minutes in a sequential order (1-2-3-4) and the sequence was repeated four times (“Cooling” sub-session; Figure 3B). The Cooling sub-session was divided into 5-minute “warm spot epochs” for analysis. The daily experimental session ended with a “Post-cooling” sub-session (free exploration at room temperature for 10 min). In addition, all mice were recorded in the home cage both before and after the experimental session (Fig. 3A). During Pre- and Post-cooling sub-sessions, the animal explored the maze relatively evenly with a moderate movement speed (Figure 3B-D), although thigmotaxis was the dominant pattern, with corners as highly preferred sites of both movement and immobility (Suppl. Fig. 2). The animals readily found the location of the warm spot after a few training sessions (median = 3). Changing the warm spot locations during Cooling induced exploration until the mouse found another warm spot and stayed on it for prolonged periods (Figure 3B and C, n = 17 session in 7 mice). Duration spent on the warm spot roughly followed a bimodal distribution with a median = 2.85 minutes (Suppl. Fig. 2A). Compared to Pre and Post sub-sessions, during Cooling, mice spent a smaller proportion of time in movement (Pre: 40 ± 19%, Post: 34 ± 16%, Cooling: 23 ± 12%, mean ± SD, defined as speed > 2.5 cm/s, n = 20 sessions from 7 mice; Figure 3D) and more time in immobility (Pre: 59 ± 19%, Post: 66 ± 16%, Cooling: 76.74 ± 12.41%, mean ± SD, defined as speed ≤ 2.5 cm/s; n = 20 sessions from 7 mice; Figure 3D). The mice spent most of the time in the corners of the ThermoMaze where heat was provided (Suppl. Fig. 2B), compared to Pre- and Post-cooling (Figure 3C). Once the heating of the Peltier element was turned off, the animal quickly left its location (median duration = 12.99 s, n = 20 sessions from 7 mice; Figure 3E) and searched for a new source of warmth. Mice increased their speed from 0 cm/s to 2.5 cm/s within 12.28 s after a warm spot was turned off (median, n = 20 sessions from 7 mice; Figure 3F) and found the new warm spot within 23.45 s (median, n = 20 sessions from 7 mice; Fig. 3G). In two additional male mice, we examined brain temperature changes during the Cooling sub-session by implanting a thermistor in the hippocampus (Suppl. Fig. 3A). In support of previous findings, we found brain state-dependent fluctuation of brain temperature (Suppl. Fig. 3B)36–38. However, cooling the environment per se did not correlate with brain temperature changes (Suppl. Fig. 3C-E), confirmation that brain temperature is strongly regulated and is largely independent of the ambient temperature38. The ThermoMaze provides an affordance for mice to select their environmental temperature through the activation of behavioral thermoregulation39.

Mice track and stay immobile on hidden warm spots in the ThermoMaze.

A) Five sub-sessions constituted a daily recording session: (1) rest epoch in the home cage, (2) pre-cooling exploration epoch (Pre), (3) Cooling, (4) post-cooling exploration epoch (Post) and (5) another rest in the home cage. B) Schematic of temperature landscape changes when the animal is in the ThermoMaze (top) and example animal trajectory (below). During Cooling, one Peltier element always provided a warm spot for the animal (four Peltier elements in the 4 corners were used in this experiment). Each Peltier element was turned on for 5 minutes in a sequential order (1-2-3-4) and the sequence was repeated four times. C) Session-averaged duration of immobility (speed ≤ 2.5 cm/s) that the animal spent at each location in the ThermoMaze; Color code: temporal duration of immobility (s); white lines divide the individual Peltier elements; n = 17 session in 7 mice). D) Cumulative distribution of animal speed in the ThermoMaze during three sub-sessions from 7 mice). Median, Kruskal–Wallis test: H = 139304.10, d.f. = 2, p < 0.001. E) Animal’s distance from the previously heated Peltier element site. F) Speed of the animal centered around warm spot transitions. G) Animal’s distance from the target warm spot as a function of time (red curve: median; time 0 = onset of heating). *p < 0.05, **p < 0.01, ***p < 0.001. In all panels, box chart displays the median, the lower and upper quartiles. (see Supplementary Table 2 for exact p values and multiple comparisons).

One of the objectives in developing the ThermoMaze was to induce immobility at several locations repeatedly and for extended time periods. To confirm that this objective was achieved, we ran control sessions with the same duration as the Cooling sub-session but at room temperature (80 minutes; Suppl. Fig. 4). Under room temperature condition (3 sessions in 3 mice), mice first explored the ThermoMaze and settled in one of the corners for an extended period of time. Although mice spent a similar total amount of time immobile under both conditions, the spatial distribution of immobility durations was more uniform in the Cooling sub-session (Suppl. Fig. 4) because the ThermoMaze paradigm forced the animals to leave their chosen spot and move to the experimenter-designated locations, i.e., the new warm spots away from the corner (Suppl. Fig. 5).

Firing rate maps of hippocampal neurons in the ThermoMaze

Compared to spatial learning and memory paradigms such as the Morris water maze40, the ThermoMaze has a non-aqueous environment and thus allows for an easy setup of electrophysiological recording. We recorded neurons from the CA1 hippocampal region by multi-shank silicon probes and separated them into putative pyramidal cells and interneurons (Methods - Unit isolation and classification section). We separated behavioral states (movement or immobility) based on movement speed (speed ≧ 2.5 cm/s = movement and speed < 2.5 cm/s = immobility).

To construct spike count maps for comparing sub-sessions, the ThermoMaze was divided into 25 x 25 bins and the number of spikes emitted by a neuron in each bin was counted and normalized by the time the mouse spent in each spatial bin. The impact of cooling during movement (theta state) was compared by calculating the correlation coefficients between Pre and Post, Pre and Cooling, and Cooling and Post spike count maps (Suppl Fig. 6A). The correlation coefficients decreased significantly across all sub-sessions, with the largest change observed between Pre-cooling and Post-cooling spike count maps in the experimental mice (Suppl Fig. 6B). Thus, the Cooling sub-session in the ThermoMaze induced a moderate decorrelation of pyramidal cells’ rate maps. Such observation constrained our ability to decode spatial information from the spiking activity during SWP-Rs in the Cooling sub-session using firing rate maps constructed during Pre- and Post-cooling sub-sessions41, because the Bayesian approaches have an underlying assumption that the spatial representation (tuning functions, or rate maps) is temporally stable.

In principle, comparison of place maps during the first and last 10 min of a 100 min session at room temperature should serve as controls. However, at room temperature mice “designate” one of the corners as home base after a few minutes of exploration and stay in that corner for the rest of the session (Suppl. Fig. 4A). Thus, exploration of the maze at the end of the session was not available.

Place-selective neuronal firing during SPW-Rs at experimenter-designated locations

As expected, SPW-Rs occurred predominantly in the corners (Fig. 4A), where the mice spent most of their time resting (Fig. 3C). Compared to room-temperature control sessions where animals spent most of their time in one corner, the spatial distribution of SPW-Rs in the Cooling sub-session was more uniform (Suppl. Fig. 4A-D), indicating that the ThermoMaze paradigm successfully biased where SPW-Rs were generated. The duration and amplitude of SPW-Rs were comparable in the ThermoMaze and the homecage (Fig. 4 B, C), whereas the mean peak frequency of SPW-Rs were significantly lower (Fig. 4D). This decrease can be explained by the lower brain temperate during sleep, a state in which the animals spent most of their time in the home cage36.

Location-specific distribution of SPW-R in the ThermoMaze

A) Spatial map of the number of SPW-Rs during the Cooling sub-session averaged across all sessions (Color code: average number of SPW-Rs per session at each location). Session-average number of SPW-Rs during Cooling was 627.3 (corresponding to 0.136 Hz). B-D) Boxplots of SPW-R properties in ThermoMaze and in the home cage (n = 19 sessions in n = 7 mice). B) Mean ripple duration in seconds (s; p = 0.108). C) Mean ripple amplitude in μV (p = 0.9). D) Mean ripple peak frequency in Hz (p < 0.001). Dots (females) and diamonds (males) of the same color represent the same animal.

To quantify spatial tuning features of neuronal firing during SPW-Rs in the ThermoMaze during the Cooling subsession, we defined a metric referred to as “spatial tuning score” (STS). We first binned the floor of the ThermoMaze into four quadrants (2x2). For each neuron, we calculated its average firing rate within SPW-Rs in each quadrant. STS was then defined by the firing rate in the quadrant with the highest within-SPW-R firing rate divided by the sum of the within-SPW-R firing rates in all four quadrants (yielding a value between 0 and 1; Fig. 5A). To test the significance of STS, we compared the STS values with their shuffled versions by randomly assigning one of the four quadrants to each SPW-R. The distribution of the STS in actual SPW-Rs was significantly higher compared to shuffled controls (Fig. 5B). Additionally, pyramidal cells exhibited higher STSs compared to interneurons (medians: pyramidal cells = 0.3432; interneurons = 0.2934; one-sided Wilcoxon rank sum test, p < 0.001). In summary, both excitatory and inhibitory neuronal populations exhibit place-selective firing during SPW-Rs, while the excitatory neurons demonstrate a stronger place-specific firing.

Spikes of CA1 pyramidal neurons during awake SPW-Rs are spatially tuned.

A) Within SPW-R firing rate maps (ThermoMaze binned into quadrants) of 6 example cells with high within SPW-R spatial tuning score (STS; from left to right, top to bottom, STS= 0.458, 0.639, 0.592, 0.672, 0.655, 0.660 respectively). Color represents within SPW-R firing rate (in Hz) of the neuron in each quadrant of the ThermoMaze. B) Cumulative distribution of spatial tuning scores of pyramidal neurons (top; n = 1150; p < 0.001) and interneurons (bottom; n = 288; p < 0.001) during SPW-Rs. Chance levels were calculated by shuffling the quadrant identity of the SPW-Rs. One-sided Wilcoxon rank sum tests. C) Bayesian decoding of the mouse’s location (quadrant of the ThermoMaze) from spike content of SPW-Rs in an example session (blue: actual ripple location; green: decoded locations; red: locations of the warm spot; session decoding accuracy = 0.65; chance level = 0.26). D) Histogram of session Bayesian decoding accuracies of ripple locations using spiking rate maps constructed during ripples as templates (with a uniform prior and a 100-fold cross-validation; P < 0.001). One-sample t-test. E) Firing rate ratios of pyramidal cells constructed during SPW-Rs and movement are positively correlated (Pearson’s r = 0.321, p < 0.001). The firing rate ratio measures the firing rate of a cell in one quadrant versus the sum of its firing rates in all four quadrants under a specific condition (within-ripple or during movement). F) Matrix of the pairwise correlation coefficient between each pair of firing rate ratio population vectors constructed during SPW-Rs and movements in different quadrants (x and y axes). Color represents Pearson’s r.

To quantify how well CA1 neurons encode spatial information during SPW-Rs at the population level, we carried out a Bayesian decoding analysis to read out the current position of the animal from spiking activity41. We constructed firing rate map templates using spikes within SWP-Rs in the training dataset and determined animal positions that maximized the likelihood of observing the spike train during SWP-Rs in the testing dataset (see Method). Spiking activity during SPW-Rs reliably identified the quadrant that the animal was in above chance level (Figure 5C, D) irrespective whether we incorporated the spatial distribution priors into the decoder in an example session (Figure 5C) or used a uniform prior (Figure 5D).

To relate spatial content of spikes during SPW-Rs and locomotion, we examined whether the same or different groups of neurons contributed to the place-specific firing during SPW-R and locomotion by calculating the firing rate ratios within preferred quadrant versus all quadrants. These ratios during SPW-Rs and movement were positively correlated (Fig. 5E; n = 1150 pyramidal cells in 20 sessions from 7 mice), suggesting that place cells42 during movement preserved their spatial properties during SPW-Rs (see also Suppl. Fig. 7 for further analysis and findings on interneurons).

Finally, we tested whether the preservation of spatial features of neuronal spiking also holds at the population level by constructing population vectors separately during movement and SPW-Rs. We then computed the pairwise correlation coefficients between these two conditions. As was the case for individual pyramidal cells, population vectors for the same quadrant during movement were similar to those during SPW-Rs (Figure 5F). Overall, these findings support and extend the observation that spiking activity during SPW-Rs continue to be influenced by the animal’s current position43.

To test specifically whether perceptual sensing of environmental features is critical in position-specific firing of neurons during SPW-Rs, we prolonged the duration of warm spots. After the Pre-cooling sub-session, the ThermoMaze temperature was decreased to 16 °C for 80 min and two Peltier elements were heated in an alternate fashion to 30 °C for 20 min (Figure 6A). As expected, mice spent most of the time immobile on the warm spots (Figure 6A,B). Similar to the 5-minute protocol (Fig. 4A), SPW-Rs occurred predominantly on the warm spots (Fig. 6C). The increased duration of stay on the warm spot facilitated the occurrence of sleep, as quantified by our brain state scoring algorithm (Fig. 6D, SPW-Rs). REM sleep was not detected since REM state typically emerges after 20-30 min of NREM episodes44. Mice spent a higher fraction of their time in sleep during the 20 min, compared to the 5 min sub-session (p = 0.003, n = 19 sessions in 7 mice and n = 7 sessions in 4 mice, Suppl. Table 1). The average inter-NREM interval was 1000 seconds (Fig. 6F, n = 7 sessions in 4 mice). Comparing the spike content of SPW-Rs during awake immobility and NREM sleep, we found that Pearson’s correlation coefficients between population vectors constructed during waking movement and waking SPW-Rs were higher than between movement and NREM SPW-Rs (Fig. 6G). These findings further support the view that sensory inputs during waking SPW-Rs can affect spiking content of SPW-Rs.

Mice sleep at experimenter-defined locations.

A) Schematic of ThermoMaze with warm spot locations (top) and the trajectory of an example animal (bottom; red rectangles correspond to the location of warm spots). During Cooling, one Peltier element was turned on for 20 min followed by another (1-2) and the sequence was repeated two times. B) Session-averaged duration of immobility (speed ≤ 2.5 cm/s) at each location in the ThermoMaze; white lines divide the individual Peltier elements (n = 7 sessions, n = 4 mice). C) Spatial distribution of SPW-R occurrences (color code: average number of SPW-Rs per session at each location, n = 7 sessions, n = 4 mice). Session-average of SPW-Rs during Cooling was 775 (corresponding to 0.16 Hz). D) Long duration of heating allowed for NREM sleep occurrence during Cooling session. Brain state changes44 are shown together with SPW-Rs (green ticks). Note that NREM sleep occurs in the second half of the 20-min warming. E) Mice spent a larger fraction of time in NREM during 20 min Cooling sub-session compared to the 5 min task variant (p = 0.003, n = 19 sessions in 7 mice and n = 7 sessions in 4 mice). (F). Mice typically spent ∼1000 seconds awake between NREM epochs. G) Box charts of Pearson’s correlation coefficients between population vectors of CA1 pyramidal neurons constructed during awake SPW-Rs, movement, and NREM SPW-Rs. Median, Kruskal–Wallis test: H = 20.7, d.f. = 2, p < 0.001 (pairwise comparison: *p = 0.037 and ***p = 1.6x10-05).

Discussion

To investigate the importance of brain state transitions in a controlled manner, we developed the ThermoMaze, a behavioral paradigm that allows for the collection of large amounts of physiological data while the animal rests at distinct experimenter-controlled locations. Since the paradigm exploits natural behavior, no training or handling is necessary. We demonstrate that mice regularly explore a cold environment until a warm spot is identified. They spend most of the time on a warm spot and even fall asleep, thus exhibiting a high degree of comfort. We exploited the long immobility epochs following exploration and showed how neurons active during hippocampal sharp wave ripples (SPW-R) replay waking experience. The ThermoMaze will allow for detailed studies of brain correlates of preparatory-consummatory transitions and open new options for studying temperature homeostasis.

Warmth-seeking homeostatic behavior

There is a renewed interest in exploiting natural learning patterns, as opposed to training animals for performing complex arbitrary signal-action associations45–52. In poikilotherm animals (species whose internal temperature varies with environmental temperature), energy homeostasis is one of the most fundamental homeostatic processes. Heat homeostasis involves multiple levels of coordination from cellular to systems, from peripheral to central53,54. To maintain core body temperature, thermogenic tissues rapidly increase glucose utilization by brown adipose tissue and shivering by skeletal muscle55,56. The hypothalamic preoptic area (POA) is regarded as the most important thermoregulatory “center” in the brain57,58. Connecting this area of research to learning, the POA is bidirectionally connected with the limbic system and multiple cortical areas which assist both online maintenance of body temperature and preparing the body for future expected changes (“allostasis”)23,59,60. These allostatic mechanisms induce exploratory behavior, searching for a warmer environment61,62. A location that provides a warm shelter needs to be remembered and generalized for future strategies. Our paradigm offers means to investigate exploratory-consummatory transitions, wake-sleep continuity in the same physical location and, in the reverse direction, the physiological processes that evaluate discomfort levels, motivate behavioral transition from rest to exploration and the circuit mechanisms that give rise to overt behaviors.

Mice, and rodents in general, are acrophobic and agoraphobic and tend to avoid open areas. Instead, they tend to move close to the wall and spend most of their non-exploration time in corners63. Thus, while we were able to train mice to seek out and stay on warm spots in the center of the maze after extensive training, their evolutionary “counter-preparedness”47 to stay in predator-prone open areas competed with the reward of warming. While these trained mice did stay transiently on the central warm spot, they spent more time returning to the corners. Our mice were on a normal day-light schedule thus their training during the day coincided with their sleep cycle. This explains why after 5-10 min spent on the safe and temperature-comfortable corner warm spots they regularly fell asleep. Yet, we noticed that mice did not simply transition from walking to immobility but, instead, even after finding the warm spot they regularly and repeatedly explored the rest of the maze before returning to the newly identified home base. By changing the temperature difference between the environment and the warm spot, it will be possible to generate psychophysical curves to quantify the competition between homeostatic and exploratory drives in future experiments. These measures, in turn, could be used to study the impact of perturbing peripheral and central energy-regulating mechanisms.

For several applications, it is not needed to tile the entire floor of the maze with Peltier elements. For example, a radial-arm maze with cooled floors or placed in a cold box can be equipped with heating Peltier elements at the ends of maze arms and center, allowing the experimenter to induce ambulation in the 1-dimensional arms, followed by extended immobility and sleep at designated areas. In a way, the ThermoMaze is analogous to the water maze40, also an avoidance task, but many more trials can be achieved in a single session and without the inconvenience of a wet environment.

SPW-R spiking content biased by current position of animal depending on brain state

We demonstrate the utility of the ThermoMaze for addressing long-standing questions in hippocampal physiology. Preparatory and consummatory behaviors in the hippocampus are associated with theta oscillations and SPW-Rs, respectively10. SPW-Rs also occur during NREM sleep but studying the differences between waking and sleep SPW-Rs has been hampered by the paucity of SPW-Rs in typical learning paradigms21,22,64–67. Neural activity during SPW-Rs has been shown to replay activity patterns observed during previous spatial navigation experiences21,43,64 and can even be predictive of activity during future experiences68–70. However, the extent to which SWP-R spiking context is biased by the current position of the animal is less known, as systematic control of position during rest/sleep has posed difficulty. The ThermoMaze enables the experimenter to control the animal’s position during SWP-R states. In agreement with previous studies43,65,67, we found that neurons whose place fields overlapped with the quadrant of the maze had a higher participation probability in SPW-Rs occurring at that location compared to other neurons. This observation supports the notion that waking replay events can be biased by perceiving features of the surrounding environment43. However, when the mouse fell asleep at the same location this relationship was weakened but did not disappear. Another potential explanation for the decreased correlation between sleep SPW-R and waking exploration is deterioration of replay as the function of time15. Alternatively, the persisting significant correlation between sleep SPW-Rs and previous exploration may also indicate that factors other than the perception of the animal’s vicinity is responsible for sleep replay70–72. Continuity of waking experience replay in waking and sleep SPW-Rs have been hypothesized previously but not yet tested73. Using the ThermoMaze, this and other related questions can now be addressed quantitatively.

Supporting information

eLife Assessment

The ThermoMaze represents a valuable tool to control the rest/exploration states of an animal. The data, collected and analyzed using solid and validated methodology, demonstrate its use in addressing previously elusive questions. More in-depth analysis of place cell activity would provide better support for some of the claims.

Reviewer #1 (Public Review):

Summary:

This manuscript introduced a new behavioral apparatus to regulate the animal's behavioral state naturally. It is a thermal maze where different sectors of the maze can be set to different temperatures; once the rest area of the animal is cooled down, it will start searching for a warmer alternative region to settle down again. They recorded with silicon probes from the hippocampus in the maze and found that the incidence of SWRs was higher at the rest areas and place cells representing a rest area were preferentially active during rest-SWRs as well but not during non-REM sleep.

Strengths:

The maze can have many future applications, e.g., see how the duration of waking immobility can influence learning, future memory recall, or sleep reactivation. It represents an out-of-the-box thinking to study and control less-studies aspects of the animals' behavior.

Weaknesses:

The impact is only within behavioral research and hippocampal electrophysiology.

Reviewer #2 (Public Review):

Summary:

In this manuscript, Vöröslakos and colleagues describe a new behavioural testing apparatus called ThermoMaze, which should facilitate controlling when a mouse is exploring the environment vs. remaining immobile. The floor of the apparatus is tiled with 25 plates, which can be individually heated, whereas the rest of the environment is cooled. The mouse avoids cooled areas and stays immobile on a heated tile. The authors systematically changed the location of the heated tile to trigger the mouse's exploratory behaviours. The authors showed that if the same plate stays heated longer, the mouse falls into an NREM sleep state. The authors conclude their apparatus allows easy control of triggering behaviours such as running/exploration, immobility and NREM sleep. The authors also carried out single-unit recordings of CA1 hippocampal cells using various silicone probes. They show that the location of a mouse can be decoded with above-chance accuracy from cell activity during sharp wave ripples, which tend to occur when the mouse is immobile or asleep. The authors suggest that consistent with some previous results, SPW-Rs encode the mouse's current location and any other information they may encode (such as past and future locations, usually associated with them).

Strengths:

Overall, the apparatus may open fruitful avenues for future research to uncover the physiology of transitions from different behavioural states such as locomotion, immobility, and sleep. The setup is compatible with neural recordings. No training is required.

Weaknesses:

I have a few concerns related to the authors' methodology and some limitations of the apparatus's current form. Although the authors suggest that switching between the plates forces animal behaviour into an exploratory mode, leading to a better sampling of the enclosure, their example position heat maps and trajectories suggest that the behaviour is still very stereotypical, restricted mostly to the trajectories along the walls or the diagonal ones (between two opposite corners). This may not be ideal for studying spatial responses known to be affected by the stereotypicity of the animal's trajectories. Moreover, given such stereotypicity of the trajectories mice take before and after reaching a specific plate, it may be that the stable activity of SWR-P ripples used for decoding different quadrants may be representing future and/or past trajectories rather than the current locations suggested by the authors. If this is the case, it may be confusing/misleading to call such activity ' place-selective firing', since they don't necessarily encode a given place per se (line 281).

Another main study limitation is the reported instability of the location cells in the Thermomaze. This may be related to the heating procedure, differences in stereotypical sampling of the enclosure, or the enclosure size (too small to properly reveal the place code). It would be helpful if the authors separate pyramidal cells into place and non-place cells to better understand how stable place cell activity is. This information may also help to disambiguate the SPW-R-related limitations outlined above and may help to solve the poor decoding problem reported by the authors (lines 218-221).

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

This manuscript introduced a new behavioral apparatus to regulate the animal's behavioral state naturally. It is a thermal maze where different sectors of the maze can be set to different temperatures; once the rest area of the animal is cooled down, it will start searching for a warmer alternative region to settle down again. They recorded with silicon probes from the hippocampus in the maze and found that the incidence of SWRs was higher at the rest areas and place cells representing a rest area were preferentially active during rest-SWRs as well but not during non-REM sleep.

We thank the reviewer for carefully reading our manuscript and providing useful and constructive comments.

Strengths:

The maze can have many future applications, e.g., see how the duration of waking immobility can influence learning, future memory recall, or sleep reactivation. It represents an out-of-the-box thinking to study and control less-studies aspects of the animals' behavior.

Weaknesses:

The impact is only within behavioral research and hippocampal electrophysiology.

We agree with this assessment but would like to add that the intersection of electrophysiological recordings in behaving animals is a very large field. Behavioral thermoregulation is a hotly researched area also by investigators using molecular tools as well. The ThermoMaze can be used for juxtacellular/intracellular recordings in behaving animals. Restricting the animal’s movement during these recordings can improve the length of recording time and recorded single unit yield in these experiments.

Moreover, the fact that animals can sleep within the task can open up new possibilities to compare the role of sleep in learning without having to move the animal from a maze back into its home cage. The cooling procedure can be easily adapted to head-fixed virtual reality experiments as well.

I have only a few questions and suggestions for future analysis if data is available.

Comment-1: Could you observe a relationship between the duration of immobility and the preferred SWR activation of place cells coding for the current (SWR) location of the animal? In the cited O'Neill et al. paper, they found that the 'spatial selectivity' of SWR activity gradually diminished within a 2-5min period, and after about 5min, SWR activity was no longer influenced by the current location of the animal. Of course, I can imagine that overall, animals are more alert here, so even over more extended immobility periods, SWRs may recruit place cells coding for the current location of the animal.

We thank the reviewer for raising this question, which is a fundamental issue that we attempted to address using the ThermoMaze. First, we indeed observed persistent place-specific firing of CA1 neurons for up to around 5 minutes, which was the maximal duration of each warm spot epoch, as shown by the decoding analysis (based on firing rate map templates constructed during SPW-Rs) in Figure 5C and D. However, we did not observe above-chance-level decoding of the current position of the animal during sharp-wave ripples using templates constructed during theta, which aligns with previous observation that CA1 neurons during “iSWRs” (15–30 s time windows surrounding theta oscillations) did not show significant differences in their peak firing rate inside versus outside the place field (O’Neil et al., 2006). We reasoned that this could be potentially explained by a different (although correlated, see Figure 5E) neuronal representation of space during theta and during awake SPW-R.

Comment-2: Following the logic above, if possible, it would be interesting to compare immobility periods on the thermal maze and the home cage beyond SWRs, as it could give further insights into differences in rest states associated with different alertness levels. E.g., power spectra may show a stronger theta band or reduced delta band compared to the home cage.

If we are correct the Reviewer would like to know whether the brain state of the animal was similar in the ThermoMaze (warm spot location) and in the home cage during immobility. A comparison of the time-evolved power spectra shows similar changes from walking to immobility in both situations without notable differences. This analysis was performed on a subset of animals (n = 17 sessions in 7 mice) that were equipped with an accelerometer (home cage behavior was not monitored by video). We detected rest epochs that lasted at least 2 seconds during wakefulness in both the home cage and ThermoMaze. Using these time points we calculated the event-triggered power spectra for the delta and theta band (±2 s around the transition time) and found no difference between the home cage and ThermoMaze (Suppl. Fig. 4D).

Prompted by the Reviewer’s question, we further quantified the changes in LFP in the two environments. We did not find any significant change in the frequencies between 1-40 Hz during Awake periods, but we did find higher delta power (1-4 Hz) in some animals in the ThermoMaze (Suppl. Fig. 4A, B).

We have also quantified the delta and theta power spectra in the few cases, when the warm spot was maintained, and the animal fell asleep. The time-resolved spectra classified the brain state as NREM, similar to sleeping in the home cage. Both delta and theta power were higher in the ThermoMaze following Awake-NREM transitions (±30 seconds around the transition, Suppl. Fig. 4C). It might well be that immobility/sleep outside the mouse’s nest might reflect some minor (but important) differences but our experiments with only a single camera recording do not have the needed resolution to reveal minor differences in posture.

We added these results to the revised Supplementary material (Suppl. Fig. 4).

Comment-3: Was there any behavioral tracking performed on naïve animals that were placed the first time in the thermal maze? I would expect some degree of learning to take place as the animal realizes that it can find another warm zone and that it is worth settling down in that area for a while. Perhaps such a learning effect could be quantified.

Unfortunately, we did not record videos during the first few sessions in the ThermoMaze. Typically, we transferred a naïve animal into the ThermoMaze for an hour on the first day to acclimatize them to the environment. This was performed without video analysis. In addition, because the current version of the maze is relatively small (20 x 20 cm), the animal usually walked around the edges of the maze before settling down at a heated warm spot. It appeared to us that there was only a very weak drive to learn the sequence and location of the warm spot, and therefore we did not quantified learning in the current experiment. We agree with the reviewer that in future studies, it will be interesting to explore whether the ThermoMaze could be adapted to a land-version of the Morris water maze by increasing the size of the maze and performing more controlled behavioral training and testing.

Comment-4: There may be a mislabeling in Figure 6g because the figure does not agree with the result text - the figure compares the population vector similarly of waking SWR vs sleep SWRs to exploration vs waking SWR and exploration vs sleep SWRs.

We thank the reviewer for raising the point, we have updated the labels accordingly.

Reviewer #2 (Public Review):

In this manuscript, Vöröslakos and colleagues describe a new behavioural testing apparatus called ThermoMaze, which should facilitate controlling when a mouse is exploring the environment vs. remaining immobile. The floor of the apparatus is tiled with 25 plates, which can be individually heated, whereas the rest of the environment is cooled. The mouse avoids cooled areas and stays immobile on a heated tile. The authors systematically changed the location of the heated tile to trigger the mouse's exploratory behaviours. The authors showed that if the same plate stays heated longer, the mouse falls into an NREM sleep state. The authors conclude their apparatus allows easy control of triggering behaviours such as running/exploration, immobility and NREM sleep. The authors also carried out single-unit recordings of CA1 hippocampal cells using various silicone probes. They show that the location of a mouse can be decoded with above-chance accuracy from cell activity during sharp wave ripples, which tend to occur when the mouse is immobile or asleep. The authors suggest that consistent with some previous results, SPW-Rs encode the mouse's current location and any other information they may encode (such as past and future locations, usually associated with them).

We thank the reviewer for carefully reading our manuscript and providing useful and constructive comments.

Strengths:

Overall, the apparatus may open fruitful avenues for future research to uncover the physiology of transitions from different behavioural states such as locomotion, immobility, and sleep. The setup is compatible with neural recordings. No training is required.

Weaknesses:

I have a few concerns related to the authors' methodology and some limitations of the apparatus's current form. Although the authors suggest that switching between the plates forces animal behaviour into an exploratory mode, leading to a better sampling of the enclosure, their example position heat maps and trajectories suggest that the behaviour is still very stereotypical, restricted mostly to the trajectories along the walls or the diagonal ones (between two opposite corners). This may not be ideal for studying spatial responses known to be affected by the stereotypicity of the animal's trajectories. Moreover, given such stereotypicity of the trajectories mice take before and after reaching a specific plate, it may be that the stable activity of SWR-P ripples used for decoding different quadrants may be representing future and/or past trajectories rather than the current locations suggested by the authors. If this is the case, it may be confusing/misleading to call such activity ' place-selective firing', since they don't necessarily encode a given place per se (line 281).

We agree with the reviewer that the current version of the ThermoMaze does not necessarily motivate the mice to sample the entire maze during warm spot transitions. However, we did show correlational evidence that neuronal firing during awake sharp-wave ripples is place-selective. Both firing rate ratios and population vectors of CA1 neurons showed a reliable correlation between those during movement and awake sharp-wave ripples (Figure 5 E and F), indicating that spatial coding during movement persists into awake SWR-P state. This finding rejects the hypothesis that neuronal firing during ripples throughout the Cooling sub-session encodes past/future trajectories, which could be explained by a lack of goal-directed behavior in order to perform the task. We hope to test whether such place-specific firing during ripples can be causally involved in maintaining an egocentric representation of space in a future study.

Besides, we have attempted to motivate the animal to visit the center of the maze during the Cooling sub-session. Moving the location of warm spots from the corners can shape the animals’ behavior and promote more exploration of the environment as we show in Suppl. Fig. 5. We agree with the Reviewer that the current size of the ThermoMaze poses these limitations. However, an example future application could be to warm the floor of a radial-arm maze by heating Peltier elements at the ends of maze arms and center in an otherwise cold room, allowing the experimenter to induce ambulation in the 1-dimensional arms, followed by extended immobility and sleep at designated areas.

Another main study limitation is the reported instability of the location cells in the Thermomaze. This may be related to the heating procedure, differences in stereotypical sampling of the enclosure, or the enclosure size (too small to properly reveal the place code). It would be helpful if the authors separate pyramidal cells into place and non-place cells to better understand how stable place cell activity is. This information may also help to disambiguate the SPW-R-related limitations outlined above and may help to solve the poor decoding problem reported by the authors (lines 218-221).

The ThermoMaze is a relatively small enclosure (20 x 20 cm) compared to typical 2D arenas (60 x 60 cm) used in hippocampal spatial studies. Due to the small environment, one possibility is that CA1 neurons encode less spatial information and only a small number of place cells could be found. Therefore, we identified place cells in each sub-session. We found 40.90%, 45.32%, and 41.26% of pyramidal cells to be place cells in the Pre-cooling, Cooling, and Post-cooling sub-sessions, respectively. Furthermore, we found on average 17.36% of pyramidal neurons pass the place cell criteria in all three sub-sessions in a daily session. Therefore, the strong decorrelation of spatial firing maps across sub-sessions cannot be explained by poor recording quality or weak neuronal encoding of spatial information but is potentially due to changes in environmental conditions.

Some additional points/queries:

Comment-1: Since the authors managed to induce sleeping on the warm pads during the prolonged stays, can they check their hypothesis that the difference in the mean ripple peak frequency (Fig. 4D) between the home cage and Thermomaze was due to the sleep vs. non-sleep states?

In response to the reviewer’s comment, we compared the ripple peak frequency that occurred during wakefulness and NREM epochs in the home cage and ThermoMaze (n = 7 sessions in 4 mice). We found that the peak frequency of the awake ripples was higher compared to both home cage and ThermoMaze NREM sleep (one-way ANOVA with Tukey’s posthoc test, ripple frequencies were: 171.63 ± 11.69, 172.21 ± 11.86, 168.19 ± 11.10 and 168.26 ± 11.08 Hz mean±SD for home cage awake, ThermoMaze awake, home cage NREM and ThermoMaze NREM conditions, p < 0.001 between awake and NREM states). We added this quantification to the revised manuscript.

NREM sleep either in home cage or in ThermoMaze affects ripple mean peak frequency similarly.

Comment-2: How many cells per mouse were recorded? How many of them were place cells? How many place cells at the same time on average? What are the place field size, peak, and mean firing rate distributions in these various conditions? It would be helpful if they could report this.

For each animal on a given day, the average number of cells recorded was 57.5, which depended on the electrodes and duration after implantation. We first applied peak firing rate and spatial information thresholds to identify place cells in each sub-session (see more details in the revised Methods section for place cell definition). We found 40.90%, 45.32%, and 41.26% of pyramidal cells to be place cells in the Pre-cooling, Cooling, and Post-cooling sub-sessions respectively. Furthermore, we found on average 17.36% of pyramidal neurons pass the place cell criteria in all three sub-sessions in a daily session.

For place cells identified in each sub-session, their place fields size is on average 61.03, 79.86, and 57.51 cm2 (standard deviation = 60.13, 69.98, and 49.64 cm2; Pre-cooling, Cooling, and Post-cooling correspondingly). A place field was defined to be a contiguous region of at least 20 cm2 (20 spatial bins) in which the firing rate was above 60% of the peak firing rate of the cell in the maze (Roux and Buzsaki et al., 2017). A place field also needs to contain at least one bin above 80% of the peak firing rate in the maze. With such definition, the average place field peak firing rate is 5.84, 5.22, and 6.48 Hz (standard deviation = 5.11, 4.65, and 5.83 Hz) and the average mean firing rate within the place fields is 4.54, 4.05, and 5.07 Hz (standard deviation = 4.00, 3.60, and 4.60).

We would like to point out that these values depend strongly on the definition of place fields, which vary widely across studies. We reason that the ThermoMaze paradigm induced place field remapping which has been reported to occur upon changes in the environment such as visual cues (Leutgeb et al., 2009). We hypothesize that temperature gradient is an important aspect among the environmental cues, thus remapping is expected. Overall, we did not aim for biological discoveries in the first presentation of the ThermoMaze. Instead, our limited goal was the detailed description of the method and its validation for behavioral and physiological experiments.

References

(1) Mizuseki K, Royer S, Diba K, Buzsáki G. Activity dynamics and behavioral correlates of CA3 and CA1 hippocampal pyramidal neurons. Hippocampus. 2012 Aug;22(8):1659-80. doi: 10.1002/hipo.22002. Epub 2012 Feb 27. PMID: 22367959; PMCID: PMC3718552.

(2) Skaggs WE,McNaughton BL,Gothard KM,Markus EJ. 1993. An information-theoretic approach to deciphering the hippocampal code. In: SJ Hanson, JD Cowan, CL Giles, editors. Advances in Neural Information Processing Systems, Vol. 5. San Francisco, CA: Morgan Kaufmann. pp 1030–1037.

(3) Roux L, Hu B, Eichler R, Stark E, Buzsáki G. Sharp wave ripples during learning stabilize the hippocampal spatial map. Nat Neurosci. 2017 Jun;20(6):845-853. doi: 10.1038/nn.4543. Epub 2017 Apr 10. PMID: 28394323; PMCID: PMC5446786.

(4) Markus, E.J., Barnes, C.A., McNaughton, B.L., Gladden, V.L. & Skaggs, W.E. Spatial information content and reliability of hippocampal CA1 neurons: effects of visual input. Hippocampus 4, 410–421 (1994).

Version 2

Abstract

Abstract

Brain states fluctuate between exploratory and consummatory phases of behavior. These state changes affect both internal computation and the organism’s responses to sensory inputs. Understanding neuronal mechanisms supporting exploratory and consummatory states and their switching requires experimental control of behavioral shifts and collecting sufficient amounts of brain data. To achieve this goal, we developed the ThermoMaze, which exploits the animal’s natural warmth-seeking homeostatic behavior. By decreasing the floor temperature and selectively heating unmarked areas, mice avoid the aversive state by exploring the maze and finding the warm spot. In its design, the ThermoMaze is analogous to the widely used water maze but without the inconvenience of a wet environment and, therefore, allows the collection of physiological data in many trials. We combined the ThermoMaze with electrophysiology recording, and report that spiking activity of hippocampal CA1 neurons during sharp-wave ripple events encode the position of the animal. Thus, place-specific firing is not confined to locomotion and associated theta oscillations but persist during waking immobility and sleep at the same location. The ThermoMaze will allow for detailed studies of brain correlates of immobility, preparatory-consummatory transitions and open new options for studying behavior-mediated temperature homeostasis.

Introduction

All behaviors can be considered as parts of a sequence of action-rest transition1. Brain states in vertebrates fall into dichotomous categories, and correspond roughly to what early behavioral research referred to as “preparatory” (or “exploratory”) and “consummatory” (or “terminal”) classes2. In mammals, these two fundamental brain states can be readily identified by basic electrophysiological monitoring of various brain structures3. They are also referred to as voluntary and non-voluntary or conscious and non-conscious brain states3. Switching between these states is correlated with high and low release of subcortical neuromodulators4–9. Consummatory/terminal behaviors include feeding and drinking, resting and its extreme form, non-rapid eye movement (NREM) sleep, while preparatory/exploratory behaviors include locomotion and other movements that are a result of a general tendency to sample environmental stimulus. Preparatory and consummatory behaviors in the hippocampus are associated with theta oscillations and sharp wave ripples (SPW-Rs), respectively10.

Deciphering the physiological underpinnings of these categories and revealing the significance of brain state transitions for cognition requires sufficient sampling of the relevant brain states. This is usually achieved by extended repeated recordings or, when possible, recording large numbers of neurons simultaneously. Prolongation of exploratory behavior can be readily achieved by placing the animal in novel environments, by food or water deprivation or by introducing delays in choice behavior tasks11,12. Recently, the honeycomb maze paradigm was introduced to extend the observation periods of exploratory deliberation13.

In contrast, the experimental control of consummatory behavioral classes is more difficult. Sleep provides an opportunity for long recordings. Comparison of sleep before and after learning is a standard paradigm to examine experience-induced brain plasticity14,15. Consummatory brain states associated with eating, drinking, and sex change rapidly with satiety and require prolonged periods of deprivation16–19. Controlling periods of awake immobility is most difficult20–22, mainly because forced immobilization of the animal is stressful23 and is accompanied by altered physiological states24.

Here we introduce the ThermoMaze, a behavioral paradigm that allows for the collection of large amounts of physiological data while the animal rests at distinct experimenter-controlled locations. In standard laboratory environments (20-24 °C)25, both housing and data collection take place below the thermoneutral zone of mice (26-34 °C)26–28. The ThermoMaze exploits the animal’s behavioral thermoregulation mechanisms29,30 and promotes thermotaxis (i.e., movement in response to environmental temperature)31. Searching for a warmer environment, social crowding and nest building are natural behavioral components of heat homeostasis31–33. The ThermoMaze allows the experimenter to guide small rodents to multiple positions in a two-dimensional environment. Decreasing the maze floor temperature induces heat-seeking behavior and after finding a warm spot, the animal stays immobile at that spot for extended periods of time, allowing for recording large amounts of neurophysiological data in immobility-related brain states. We report on both behavioral control and hippocampal electrophysiological correlates of heat-seeking activity to illustrate the versatile utility of the ThermoMaze.

Results

Design and Construction of the ThermoMaze

The ThermoMaze is designed to guide small rodents to warm spatial locations in a two-dimensional cold environment, consisting of a box (width, length, height: 20, 20, 40 cm, respectively) made from an acrylic plexiglass sheet (Fig. 1A, top). The floor of the maze is constructed from 25 Peltier elements (40 x 40 x 3.6 mm) that are attached to aluminum water cooling block heatsinks (40 x 40 x 12 mm, n = 25) with heat-conductive epoxy and are insulated from each other by wood epoxy (Fig. 1A, dashed inset). Each Peltier element is controlled by an electrically operated switch (relay) that opens and closes high-current circuits by receiving transistor-transistor-logic (TTL) signals from outside sources (Fig. 1B). Peltier elements can be heated individually up to 30 °C to provide a “warm spot” for the animal when other regions of the floor are under cooling (Fig. 1B, active heating of one Peltier element is shown). The ambient temperature of the maze is controlled by water circulated from the water tank through the water-cooling blocks. We set the floor temperature to either ∼25°C (room temperature) or to ∼10 °C (cooling, Fig. 1C and Suppl. Fig. 1), but a range of ambient temperatures (5-30 °C) could be employed. The water temperature is monitored by a K-type thermocouple placed inside the water tank (Fig. 1A bottom). The floor temperature of the ThermoMaze is monitored using a thermal camera (FLIR C5) providing continuous registration of real-time temperature changes (Fig. 1A).

Construction and temperature control of the ThermoMaze.

A) Schematic of the ThermoMaze. The floor was built using 25 Peltier elements attached to water cooling block heatsinks (building block). The position of the animal and the temperature of the ThermoMaze can be recorded using a video camera and an infrared camera positioned above the box, respectively. An ‘X’ was taped inside the maze as an external cue below the camera synchronizing LED. Water circulates through the water cooling heatsinks using a water pump submerged in a water tank (one row of heatsinks is attached to one pump). The temperature of the water tank is monitored and recorded using a thermocouple (white symbol inside water tank, DAQ – analog input of the data acquisition system). Peltier elements are connected to a power supply (red and blue dots represent the anode and cathode connection). B) Circuit diagram and schematic of Peltier elements (n = 25), viewed from the top. TTL pulses generated by an AVR-based microcontroller board (Arduino Mega 2560) close a relay switch connected to a variable voltage power source. Each Peltier element can be independently heated (surface temperature depends on applied voltage and temperature difference between hot and cold plate of Peltier element). C) Schematic of the water circulation cooling system, viewed from the bottom of the floor (each Peltier element has its own water-cooling aluminum heatsink, shown in silver, n = 25). Five submerging DC pumps are used to circulate water across 25 heatsinks (dashed lines show the Peltier elements connected to one pump). The temperature of the heatsink is transferred to the Peltier element passively through the silver epoxy resulting in passive cooling of the floor of the ThermoMaze.

Prior to the experiments, the thermal camera, which continuously measures the surface temperature of the floor of the ThermoMaze is calibrated by thermocouples placed directly on Peltier elements (Fig. 2). The accuracy of the FLIR 5C infrared camera is ±3 °C. With proper calibration and attention to emissivity (an object’s ability to emit rather than reflect infrared energy) the margin of error can be less than 1 °C34.

Calibration of the ThermoMaze temperature regulation.

A) Side view of the ThermoMaze. Prior to animal experiments, we calibrated the heating and cooling performance of the Peltier elements and temperature measurement. We attached thermocouples (white symbols) to the surface of the Peltier elements serving as the ground-truth for calibrating the infrared camera placed above the ThermoMaze. Different voltage levels were used for the calibration (2.2, 2.4, 2.6, 2.8 and 3V) while the water tank temperature was kept constant. B) Top: four Peltier elements used in later experiments are chosen for calibration (four corners). Bottom: one minute heating was repeated four times at each voltage level. C) Simultaneously recorded temperature by thermocouples (left) and infrared camera (right). Increasing voltages induced increased heating (n = 4 trials per intensity, mean ± SD are shown). While the temporal dynamics yielded similar results between the two systems, we found ∼4 °C offset between infrared and thermocouple-measured signals. D) Temperature changes of four Peltier elements used during an emulated behavioral session (without any animal subject) tracked by thermocouples. E) Temporal dynamics of temperature changes at the four Peltier elements during active heating and following passive cooling. The temperature reaches steady state within 31 ± 10.3 seconds (mean ± SD, n = 4 trials across 4 Peltier elements).

Mice seek out hidden warm spots in the ThermoMaze

To illustrate the novel advantages of the ThermoMaze on behavior and brain activity, we tested 11 mice (n = 3 male and 8 female mice) with silicon probe recordings from the hippocampus (Suppl. Table 1). One wall was marked by a prominent visual cue (black tape and blinking light-emitting diode; LED) to provide a distinct spatial cue in the box (Fig. 1A)35. On each experimental day, the mouse was placed in the ThermoMaze and allowed to explore it for 10 min at room temperature (“Pre-cooling” sub-session; Figure 3A). Next, the ThermoMaze temperature was decreased to around 14 °C for 80 min and four Peltier elements (“warm spots”; typically, in the corners) were sequentially and repeatedly turned on and heated up to 30 °C. One Peltier element was turned on for 5 minutes in a sequential order (1-2-3-4) and the sequence was repeated four times (“Cooling” sub-session; Figure 3B). The Cooling sub-session was divided into 5-minute “warm spot epochs” for analysis. The daily experimental session ended with a “Post-cooling” sub-session (free exploration at room temperature for 10 min). In addition, all mice were recorded in the home cage both before and after the experimental session (Fig. 3A). During Pre- and Post-cooling sub-sessions, the animal explored the maze relatively evenly with a moderate movement speed (Figure 3B-D), although thigmotaxis was the dominant pattern, with corners as highly preferred sites of both movement and immobility (Suppl. Fig. 2). The animals readily found the location of the warm spot after a few training sessions (median = 3). Changing the warm spot locations during Cooling induced locomotion until the mouse found another warm spot and stayed on it for prolonged periods (Figure 3B and C, n = 17 sessions in 7 mice). Duration spent on the warm spot roughly followed a bimodal distribution with a median = 2.85 minutes (Suppl. Fig. 2A). Compared to Pre and Post sub-sessions, during Cooling, mice spent a smaller proportion of time in locomotion (Pre: 40 ± 19%, Post: 34 ± 16%, Cooling: 23 ± 12%, mean ± SD, defined as speed > 2.5 cm/s, n = 20 sessions from 7 mice; Figure 3D) and more time in immobility (Pre: 59 ± 19%, Post: 66 ± 16%, Cooling: 76.74 ± 12.41%, mean ± SD, defined as speed ≤ 2.5 cm/s; n = 20 sessions from 7 mice; Figure 3D). The mice spent most of the time in the corners of the ThermoMaze where heat was provided (Suppl. Fig. 2B), compared to Pre- and Post-cooling (Figure 3C). Once the heating of the Peltier element was turned off, the animal quickly left the warm spot (median duration = 12.99 s, n = 20 sessions from 7 mice; Figure 3E) and searched for a new source of warmth. Mice increased their speed from 0 cm/s to 2.5 cm/s within 12.28 s after a warm spot was turned off (median, n = 20 sessions from 7 mice; Figure 3F) and found the new warm spot within 23.45 s (median, n = 20 sessions from 7 mice; Fig. 3G). In two additional male mice, we examined brain temperature changes during the Cooling sub-session by implanting a thermistor in the hippocampus (Suppl. Fig. 3A). In support of previous findings, we found brain state-dependent fluctuation of brain temperature (Suppl. Fig. 3B)36–38. However, cooling the environment per se did not correlate with brain temperature changes (Suppl. Fig. 3C-E), confirmation that brain temperature is strongly regulated and is largely independent of the ambient temperature38. The ThermoMaze provides an affordance for mice to select their environmental temperature through the activation of behavioral thermoregulation39. We also quantified the changes in LFP (1-40 Hz) to test whether the brain state of the animal was similar in the ThermoMaze and home cage during wakeful periods. We did not find significant changes in these frequencies (Suppl. Fig. 4).

Mice track and stay immobile on hidden warm spots in the ThermoMaze.

A) Five sub-sessions constituted a daily recording session: (1) rest epoch in the home cage, (2) pre-cooling exploration epoch (Pre), (3) Cooling, (4) post-cooling exploration epoch (Post) and (5) another rest in the home cage. B) Schematic of temperature landscape changes when the animal is in the ThermoMaze (top) and example animal trajectory (below). During Cooling, one Peltier element always provided a warm spot for the animal (four Peltier elements in the 4 corners were used in this experiment). Each Peltier element was turned on for 5 minutes in a sequential order (1-2-3-4) and the sequence was repeated four times. C) Session-averaged duration of immobility (speed ≤ 2.5 cm/s) that the animal spent at each location in the ThermoMaze; Color code: temporal duration of immobility (s); white lines divide the individual Peltier elements; n = 17 session in 7 mice). D) Cumulative distribution of animal speed in the ThermoMaze during three sub-sessions from 7 mice). Median, Kruskal–Wallis test: H = 139304.10, d.f. = 2, p < 0.001. E) Animal’s distance from the previously heated Peltier element site. F) Speed of the animal centered around warm spot transitions. G) Animal’s distance from the target warm spot as a function of time (red curve: median; time 0 = onset of heating). *p < 0.05, **p < 0.01, ***p < 0.001. In all panels, box chart displays the median, the lower and upper quartiles. (see Supplementary Table 2 for exact p values and multiple comparisons).

One of the objectives in developing the ThermoMaze was to induce immobility at several locations repeatedly and for extended periods. To confirm that this objective was achieved, we ran control sessions with the same duration as the Cooling sub-session but at room temperature (80 minutes; Suppl. Fig. 5). Under room temperature conditions (3 sessions in 3 mice), mice first explored the ThermoMaze and settled in one of the corners for an extended period. Although mice spent a similar total amount of time immobile under both conditions, the spatial distribution of immobility durations was more uniform in the Cooling sub-session (Suppl. Fig. 5) because the ThermoMaze paradigm forced the animals to leave their chosen spot and move to the experimenter-designated locations, i.e., the new warm spots away from the corner (Suppl. Fig. 6).

Firing rate maps of hippocampal neurons in the ThermoMaze

Compared to spatial learning and memory paradigms such as the Morris water maze40, the ThermoMaze has a non-aqueous environment and thus allows for an easy setup of electrophysiological recording. We recorded neurons from the CA1 hippocampal region by multi-shank silicon probes and separated them into putative pyramidal cells and interneurons (Methods - Unit isolation and classification section). We separated behavioral states (movement or immobility) based on movement speed (speed ≧ 2.5 cm/s = movement and speed < 2.5 cm/s = immobility).

To construct spike count maps for comparing sub-sessions, the ThermoMaze was divided into 25 x 25 bins and the number of spikes emitted by a neuron in each bin was counted and normalized by the time the mouse spent in each spatial bin. The impact of cooling during movement (theta state) was compared by calculating the correlation coefficients between Pre and Post, Pre and Cooling, and Cooling and Post spike count maps (Suppl Fig. 7A). The correlation coefficients decreased significantly across all sub-sessions, with the largest change observed between Pre-cooling and Post-cooling spike count maps in the experimental mice (Suppl Fig. 7B). Thus, the Cooling sub-session in the ThermoMaze induced a moderate decorrelation of pyramidal cells’ rate maps, potentially explained by place field remapping upon changing environmental sensory cues (here, temperature gradient)41. Such observation constrained our ability to decode spatial information from the spiking activity during SWP-Rs in the Cooling sub-session using firing rate maps constructed during Pre- and Post-cooling sub-sessions42, because the Bayesian decoding approaches have an underlying assumption that the spatial representation (tuning functions, or rate maps) is temporally stable.

Because the ThermoMaze is a relatively small enclosure, one possibility is that CA1 neurons encode less spatial information and only a small number of place cells could be found. Therefore, we identified place cells in each sub-session separately (see more details in the Method section for place cell definition, Suppl. Fig. 8). We found 40.90%, 45.32%, and 41.26% of pyramidal cells to have place fields in the Pre-cooling, Cooling, and Post-cooling sub-sessions respectively. Furthermore, we found that, on average, 17.36% of pyramidal neurons passed the place field criteria in all three sub-sessions in a daily session. Therefore, the decorrelation of spatial firing maps across sub-sessions cannot be explained by poor recording quality or weak neuronal encoding of spatial information but is potentially due to changes in environmental cues.

In principle, comparison of place maps during the first and last 10 min of a 100 min session at room temperature should serve as controls. However, at room temperature mice “designate” one of the corners as home base after a few minutes of exploration and stay in that corner for the rest of the session (Suppl. Fig. 5A). Thus, exploration of the maze at the end of the session was not available.

Place-selective neuronal firing during SPW-Rs at experimenter-designated locations

As expected, SPW-Rs occurred predominantly in the corners (Fig. 4A), where the mice spent most of their time resting (Fig. 3C). Compared to room-temperature control sessions where animals spent most of their time in one corner, the spatial distribution of SPW-Rs in the Cooling sub-session was more uniform (Suppl. Fig. 5A-D), indicating that the ThermoMaze paradigm successfully biased where SPW-Rs were generated. The duration and amplitude of SPW-Rs were comparable in the ThermoMaze and the home cage (Fig. 4 B, C), whereas the mean peak frequency of SPW-Rs were significantly lower (Fig. 4D). This decrease can be explained by the lower brain temperate during sleep, a state in which the animals spent most of their time in the home cage36.

Location-specific distribution of SPW-R in the ThermoMaze

A) Spatial map of the number of SPW-Rs during the Cooling sub-session averaged across all sessions (Color code: average number of SPW-Rs per session at each location). Session-average number of SPW-Rs during Cooling was 627.3 (corresponding to 0.136 Hz). B-D) Boxplots of SPW-R properties in ThermoMaze and in the home cage (n = 19 sessions in n = 7 mice). B) Mean ripple duration in seconds (s; p = 0.108). C) Mean ripple amplitude in μV (p = 0.9). D) Mean ripple peak frequency in Hz (p < 0.001). Dots (females) and diamonds (males) of the same color represent the same animal.

To quantify spatial tuning features of neuronal firing during SPW-Rs in the ThermoMaze during the Cooling subsession, we defined a metric referred to as “spatial tuning score” (STS). We first binned the floor of the ThermoMaze into four quadrants (2x2). For each neuron, we calculated its average firing rate within SPW-Rs in each quadrant. STS was then defined by the firing rate in the quadrant with the highest within-SPW-R firing rate divided by the sum of the within-SPW-R firing rates in all four quadrants (yielding a value between 0 and 1; Fig. 5A). To test the significance of STS, we compared the STS values with their shuffled versions by randomly assigning one of the four quadrants to each SPW-R. The distribution of the STS in actual SPW-Rs was significantly higher compared to shuffled controls (Fig. 5B). Additionally, pyramidal cells exhibited higher STSs compared to interneurons (medians: pyramidal cells = 0.3432; interneurons = 0.2934; one-sided Wilcoxon rank sum test, p < 0.001). In summary, both excitatory and inhibitory neuronal populations exhibit place-selective firing during SPW-Rs, while the excitatory neurons demonstrate a stronger place-specific firing.

Spikes of CA1 pyramidal neurons during awake SPW-Rs are spatially tuned.

A) Within SPW-R firing rate maps (ThermoMaze divided into four quadrants) of 6 example cells with high within SPW-R spatial tuning score (STS; from left to right, top to bottom, STS= 0.458, 0.639, 0.592, 0.672, 0.655, 0.660 respectively). Color represents within SPW-R firing rate (in Hz) of the neuron in each quadrant of the ThermoMaze. B) Cumulative distribution of spatial tuning scores of pyramidal neurons (top; n = 1150; p < 0.001) and interneurons (bottom; n = 288; p < 0.001) during SPW-Rs. Chance levels were calculated by shuffling the quadrant identity of the SPW-Rs. One-sided Wilcoxon rank sum tests. C) Bayesian decoding of the mouse’s location (quadrant of the ThermoMaze) from spike content of SPW-Rs in an example session (blue: actual ripple location; green: decoded locations; red: locations of the warm spot; session decoding accuracy = 0.65; chance level = 0.26). D) Histogram of session Bayesian decoding accuracies of ripple locations using spiking rate maps constructed during ripples as templates (with a uniform prior and a 100-fold cross-validation; P < 0.001). One-sample t-test. E) Firing rate ratios of pyramidal cells constructed during SPW-Rs and movement are positively correlated (Pearson’s r = 0.321, p < 0.001). The firing rate ratio measures the firing rate of a cell in one quadrant versus the sum of its firing rates in all four quadrants under a specific condition (within-ripple or during movement). F) Matrix of the pairwise correlation coefficient between each pair of firing rate ratio population vectors constructed during SPW-Rs and movements in different quadrants (x and y axes). Color represents Pearson’s r.

To quantify how well CA1 neurons encode spatial information during SPW-Rs at the population level, we carried out a Bayesian decoding analysis to read out the current position of the animal from spiking activity42. We constructed firing rate map templates using spikes within SPW-Rs in the training dataset and determined animal positions that maximized the likelihood of observing the spike train during SPW-Rs in the testing dataset (see Method). Spiking activity during SPW-Rs reliably identified the quadrant that the animal was in above chance level (Figure 5C, D) irrespective of whether we incorporated the spatial distribution priors into the decoder in an example session (Figure 5C) or used a uniform prior (Figure 5D).

To relate the spatial content of spikes during SPW-Rs and locomotion, we examined whether the same or different groups of neurons contributed to the place-specific firing during SPW-R and locomotion by calculating the firing rate ratios within the preferred quadrant versus all quadrants. These ratios during SPW-Rs and movement were positively correlated (Fig. 5E; n = 1150 pyramidal cells in 20 sessions from 7 mice), suggesting that place cells43 during movement preserved their spatial properties during SPW-Rs (see also Suppl. Fig. 9 for further analysis and findings on interneurons).

Finally, we tested whether the preservation of spatial features of neuronal spiking also holds at the population level by constructing population vectors separately during movement and SPW-Rs. We then computed the pairwise correlation coefficients between these two conditions. As was the case for individual pyramidal cells, population vectors for the same quadrant during movement were similar to those during SPW-Rs (Figure 5F). Overall, these findings support and extend the observation that spiking activity during SPW-Rs continues to be influenced by the animal’s current position44.

To test specifically whether perceptual sensing of environmental features is critical in position-specific firing of neurons during SPW-Rs, we prolonged the duration of warm spots. After the Pre-cooling sub-session, the ThermoMaze temperature was decreased to 16 °C for 80 min and two Peltier elements were heated alternately to 30 °C for 20 min (Figure 6A). As expected, mice spent most of the time immobile on the warm spots (Figure 6A,B). Similar to the 5-minute protocol (Fig. 4A), SPW-Rs occurred predominantly on the warm spots (Fig. 6C). The increased duration of stay on the warm spot facilitated the occurrence of sleep, as quantified by our brain state scoring algorithm (Fig. 6D, SPW-Rs). REM sleep was not detected since REM state typically emerges after 20-30 min of NREM episodes45. Mice spent a higher fraction of their time in sleep during the 20 min, compared to the 5 min sub-session (p = 0.003, n = 19 sessions in 7 mice and n = 7 sessions in 4 mice, Suppl. Table 1). The average inter-NREM interval was 1000 seconds (Fig. 6F, n = 7 sessions in 4 mice). Comparing the population vector similarity of waking SPW-Rs vs. NREM SPW-Rs to movement vs. waking SPW-Rs and movement vs. NREM SPW-Rs, we found that place-specific coding during SPW-Rs persists into sleep, and we observed the highest correlation when comparing SPW-Rs during different brain states than when comparing SPW-Rs and movement (Fig. 6G). These findings further support the view that sensory inputs during waking SPW-Rs can affect the spiking content of SPW-Rs.

Mice sleep at experimenter-defined locations.

A) Schematic of ThermoMaze with warm spot locations (top) and the trajectory of an example animal (bottom; red rectangles correspond to the location of warm spots). During Cooling, one Peltier element was turned on for 20 min followed by another (1-2) and the sequence was repeated two times. B) Session-averaged duration of immobility (speed ≤ 2.5 cm/s) at each location in the ThermoMaze; white lines divide the individual Peltier elements (n = 7 sessions, n = 4 mice). C) Spatial distribution of SPW-R occurrences (color code: average number of SPW-Rs per session at each location, n = 7 sessions, n = 4 mice). Session-average of SPW-Rs during Cooling was 775 (corresponding to 0.16 Hz). D) Long duration of heating allowed for NREM sleep occurrence during Cooling sub-session in an example session. Brain state changes45 are shown together with SPW-Rs (green ticks). Note that NREM sleep occurs in the second half of the 20-min warming. E) Mice spent a larger fraction of time in NREM during 20 min Cooling sub-session compared to the 5 min task variant (p = 0.003, n = 19 sessions in 7 mice and n = 7 sessions in 4 mice). (F). Mice typically spent ∼1000 seconds awake between NREM epochs. G) Box charts of Pearson’s correlation coefficients between population vectors of CA1 pyramidal neurons constructed during awake SPW-Rs, movement, and NREM SPW-Rs. Median, Kruskal–Wallis test: H = 20.7, d.f. = 2, p < 0.001 (pairwise comparison: *p = 0.037 and ***p = 1.6x10-05).

Since sleep occurred on the warm spots during the prolonged stays, we also tested our hypothesis that the difference in the mean ripple peak frequency (Fig. 4D) between the home cage and ThermoMaze was due to the sleep vs. non-sleep states. We compared the ripple peak frequency that occurred during wakefulness and NREM epochs in the home cage and ThermoMaze (n = 7 sessions in 4 mice). We found that the peak frequency of the awake ripples was higher compared to both home cage and ThermoMaze NREM sleep (one-way ANOVA with Tukey’s posthoc test; ripple frequencies were: 171.63 ± 11.69, 172.21 ± 11.86, 168.19 ± 11.10 and 168.26 ± 11.08 Hz, mean±SD for home cage awake, ThermoMaze awake, home cage NREM and ThermoMaze NREM conditions, p < 0.001 between awake and NREM states).

Discussion

To investigate the importance of brain state transitions in a controlled manner, we developed the ThermoMaze, a behavioral paradigm that allows for the collection of large amounts of physiological data while the animal rests at distinct experimenter-controlled locations. Since the paradigm exploits natural behavior, no extensive training or handling is necessary. We demonstrate that mice regularly explore a cold environment until a warm spot is identified. They spend most of the time in a warm spot and even fall asleep, thus exhibiting a high degree of comfort. We exploited the long immobility epochs following exploration and showed how neurons active during hippocampal sharp wave ripples (SPW-R) replay waking experience. The ThermoMaze will allow for detailed studies of brain correlates of preparatory-consummatory transitions and open new options for studying temperature homeostasis.

Warmth-seeking homeostatic behavior

There is a renewed interest in exploiting natural learning patterns, as opposed to training animals for performing complex arbitrary signal-action associations46–53. In poikilotherm animals (species whose internal temperature varies with environmental temperature), energy homeostasis is one of the most fundamental homeostatic processes. Heat homeostasis involves multiple levels of coordination from cellular to systems, from peripheral to central54,55. To maintain core body temperature, thermogenic tissues rapidly increase glucose utilization by brown adipose tissue and shivering by skeletal muscle56,57. The hypothalamic preoptic area (POA) is regarded as the most important thermoregulatory “center” in the brain58,59. Connecting this area of research to learning, the POA is bidirectionally connected with the limbic system and multiple cortical areas which assist both online maintenance of body temperature and preparing the body for future expected changes (“allostasis”)23,60,61. These allostatic mechanisms induce exploratory behavior, searching for a warmer environment62,63. A location that provides a warm shelter needs to be remembered and generalized for future strategies. Our paradigm offers means to investigate exploratory-consummatory transitions, wake-sleep continuity in the same physical location and, in the reverse direction, the physiological processes that evaluate discomfort levels, motivate behavioral transition from rest to exploration, and the circuit mechanisms that give rise to overt behaviors.

Mice, and rodents in general, are acrophobic and agoraphobic and tend to avoid open areas. Instead, they tend to move close to the wall and spend most of their non-exploration time in corners64. Thus, while we were able to train mice to seek out and stay in warm spots in the center of the maze after extensive training, their evolutionary “counter-preparedness”48 to stay in predator-prone open areas competed with the reward of warming. While these trained mice did stay transiently in the central warm spot, they spent more time returning to the corners. Our mice were on a normal day-light schedule thus their training during the day coincided with their sleep cycle. This explains why after 5-10 min spent on the safe and temperature-comfortable corner warm spots they regularly fell asleep. Yet, we noticed that mice did not simply transition from walking to immobility but, instead, even after finding the warm spot they regularly and repeatedly explored the rest of the maze before returning to the newly identified home base. By changing the temperature difference between the environment and the warm spot, it will be possible to generate psychophysical curves to quantify the competition between homeostatic and exploratory drives in future experiments. These measures, in turn, could be used to study the impact of perturbing peripheral and central energy-regulating mechanisms.

For several applications, there is no need to tile the entire floor of the maze with Peltier elements. For example, a radial-arm maze with cooled floors or placed in a cold box can be equipped with heating Peltier elements at the ends of maze arms and center, allowing the experimenter to induce ambulation in the 1-dimensional arms, followed by extended immobility and sleep at designated areas. In a way, the ThermoMaze is analogous to the water maze40, also an avoidance task, but many more trials can be achieved in a single session and without the inconvenience of a wet environment.

SPW-R spiking content is biased by current position of the animal and depends on brain state

We demonstrate the utility of the ThermoMaze for addressing long-standing questions in hippocampal physiology. Preparatory and consummatory behaviors in the hippocampus are associated with theta oscillations and SPW-Rs, respectively10. SPW-Rs also occur during NREM sleep but studying the differences between waking and sleep SPW-Rs has been hampered by the paucity of SPW-Rs in typical learning paradigms21,22,65–68. Neural activity during SPW-Rs has been shown to replay activity patterns observed during previous spatial navigation experiences21,44,65 and can even be predictive of activity during future experiences69–71. However, the extent to which SWP-R spiking context is biased by the current position of the animal is less known, as systematic control of position during rest/sleep has posed difficulty. The ThermoMaze enables the experimenter to control the animal’s position during SWP-R states. In agreement with previous studies44,66,68, we found that neurons whose place fields overlapped with the quadrant of the maze had a higher participation probability in SPW-Rs occurring at that location compared to other neurons. This observation supports the notion that waking replay events can be biased by perceiving features of the surrounding environment44. However, when the mouse fell asleep at the same location this relationship was weakened but did not disappear. Another potential explanation for the decreased correlation between sleep SPW-R and waking exploration is the deterioration of replay as the function of time15. Alternatively, the persisting significant correlation between sleep SPW-Rs and previous exploration may also indicate that factors other than the perception of the animal’s vicinity are responsible for sleep replay71–73. Continuity of waking experience replay in waking and sleep SPW-Rs have been hypothesized previously but not yet tested74. Using the ThermoMaze, this and other related questions can now be addressed quantitatively.

Supporting information